---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r srr, eval = FALSE, echo = FALSE}
#' @srrstatsTODO {G1.2} *Statistical Software should include a* Life Cycle Statement *describing current and anticipated future states of development.*
#' @srrstatsTODO {G1.1} *Statistical Software should document whether the algorithm(s) it implements are:* - *The first implementation of a novel algorithm*; or - *The first implementation within **R** of an algorithm which has previously been implemented in other languages or contexts*; or - *An improvement on other implementations of similar algorithms in **R***.
```


# GLMMcosinor

<!-- badges: start -->
[![R-CMD-check](https://github.com/RWParsons/GLMMcosinor/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/RWParsons/GLMMcosinor/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/RWParsons/GLMMcosinor/branch/main/graph/badge.svg)](https://app.codecov.io/gh/RWParsons/GLMMcosinor?branch=main)
<!-- badges: end -->


The goal of GLMMcosinor is to fit a generalised linear (mixed-) model (GLM) to circadian data that may be sourced but passing on the flexibility of the glmmTMB framework. For example, this includes the ability to use link functions appropriate for outcome variables that are gaussian, poisson, gamma, or binomial. 

Existing statistical software for circadian data analyses (including `cosinor` or `circacompare`) allow the user to fit such data to a linear model, but many are limited in due to this; GLMMcosinor aims to be comprehensive and flexible. Hence, GLMMcosinor is an improvement of other implementations of similar algorithms in R. See table below for features available within different currently available software.

#### TODO add table


## Installation

You can install the development version of GLMMcosinor from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("RWParsons/GLMMcosinor")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(GLMMcosinor)
```


```{r}
data(vitamind)
model <- cosinor.glmm(Y ~ X + amp.acro(time, group = "X"), data = vitamind)
summary(model)
```

```{r, eval=FALSE}
plot(model, superimpose.data = TRUE)
polar_plot(model)
```

```{r}
p1 <- plot(model, superimpose.data = TRUE)
p2 <- polar_plot(model)
cowplot::plot_grid(p1, p2, nrow=1)
```





## Citation

```{r}
citation("GLMMcosinor")
```






