---
title: "GLMMcosinor"
author: "Oliver Jayasinghe and Rex Parsons"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction-to-GLMMcosinor}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(GLMMcosinor)
```

## Introduction

The GLMMcosinor package generates generalised linear models based on
circadian data, and allows users to summarise, predict, and plot based
on this modelling. Existing packages have focussed primarily on Gaussian
data. Some circadian regression modelling packages haveallowed users to
specify generalised linear models, but with limited flexibility.
GLMMcosinor takes a comprehensive approach to modelling by utilising the
glmmTMB package, which allows users to model circadian data from the
following distributions: - binomial - guassian - Gamma -
inverse.gaussian - poisson - quasi - quasibinomial - quasipoisson

## Using: 'cosinor.glmm.R'

cosinor.glmm() generates a regression model with parameter estimates
based on a data set and formula specified by the user.

The formula argument for cosinor.glmm() is constructed in two parts:

Firstly: specification of any intercept or covariate mesor terms. By
default, an intercept term is included but this can be removed by
including a 0 in the formula. For example, the following formula has an
intercept and X listed as a covariate mesor term:

```{r, message=F, warning=F}
data(vitamind)
cosinor.glmm(Y ~ X + amp.acro(time, group = "X"), data = vitamind)
```

Alternatively, the following example will have no intercept term, but
will still generate a mesor estimate for the X group. In the output, the
[X=0] and [X=1] refer to mesors for the levels 0 and 1 respectively
within the group "X".

```{r, message=F, warning=F}
cosinor.glmm(Y ~ 0 + X + amp.acro(time, group = "X"), data = vitamind)
```

Finally, the following will have no intercept term, nor will it generate
any mesor estimates:

```{r, message=F, warning=F}
cosinor.glmm(Y ~ 0 + amp.acro(time, group = "X"), data = vitamind)
```

Secondly, specification of the amp.acro() function. As will be described
later, the amp.acro() function modifies the original dataset so that the
amplitude and acrophase can be calculated. Importantly, the amp.acro()
argument is where users specify:

-   'group' (the name of the group covariates). Ensure that the names in
    this argument are actually found in the dataset

-   'time' (the name of the time column). Again, ensure that the name
    matches the name of the time values in the dataset. Also, ensure
    that this argument is NOT a string.

-   'n_components' (The number of components). If the user wishes to
    have a more than one trigonometric component in the model, they can
    specify the number of such components here. Note that this value
    will need to match the length of the group argument and the period
    argument. For example, if n_components = 3, then the group argument
    should have 3 characters, where the position of each character
    corresponds to the component that the group covariate describes:\

```{r, message=F, warning=F}
data(vitamind)
vitamind$Z <- rbinom(length(vitamind$X), 1, prob = 0.5)
cosinor.glmm(Y ~ X + amp.acro(time, n_components = 3, group = c("X", NA, "Z"), period = c(12, 10, 8)), data = vitamind)
```

As demonstrated in the example above, the 'NA' entry in the group
argument indicates a component with no group covariate assigned.

If only one group argument and and/or one period argument is supplied,
but there is more than one component specified, then it will be assumed
that the one group and one period specified applies to all components:

For instance, the following two cosinor.glmm() calls produce the same
output:

```{r, message=F, warning=F}
cosinor.glmm(Y ~ X + amp.acro(time, n_components = 3, group = "X", period = c(12)), data = vitamind)
cosinor.glmm(Y ~ X + amp.acro(time, n_components = 3, group = c("X", "X", "X"), period = c(12, 12, 12)), data = vitamind)
```

### Dispersion and zero-inflation model specification

The cosinor.glmm() function optionally allows users to specify formulas
for dispersion and zero-inflation models. These formulas are independent
of the main formula specification:

```{r, message=F, warning=F}
data(vitamind) # load vitamind dataset
vitamind$Z <- rbinom(length(vitamind$X), 2, prob = 0.5) # modify dataset

cosinor.glmm(Y ~ X + amp.acro(time, n_components = 3, group = c("X", NA, "Z"), period = c(12, 10, 8)),
  data = vitamind,
  dispformula = ~ X + amp.acro(time, n_components = 3, group = c("Z", NA, "X"), period = c(12, 11, 8)),
  ziformula = ~ X + amp.acro(time, n_components = 2, group = c("Z", "X"), period = c(12, 8))
)
```

This example shows how dispformula and ziformula can be specified with a
different number of components, with different periods, and with
different group-component assignments. By default, dispformula = \~1,
and ziformula = \~0

Mention: - how to specify the input (specifically, amp.acro) - how to
specify multi-component models - how to specify dispersion and
zero-inflation models - how to specify random effects - how to interpret
the output

## Using: 'summary.cosinor.glmm.R'

The summary.cosinor.glmm() function provides summary statistics for an
object produced by the cosinor.glmm() function. It outputs the
estimates, confidence intervals, and tests for both the raw parameters
and the transformed (mesor, amplitude, and acrophase) parameters.
Currently, the summary statistics do not represent a comparison between
any groups - that is the role of the test_cosinor() function.

Here is an example of how to use the summary.cosinor.glmm() function:

```{r, message=F, warning=F}
data(vitamind)
object <- cosinor.glmm(Y ~ X + amp.acro(time, group = "X"), data = vitamind)
summary(object)
```

The summary statistics for dispersion and zero-inflation models will
also be provided by the summary.cosinor.glmm() function, if the original
cosinor.glmm object being analysed contains them:

```{r, message=F, warning=F}
data(vitamind) # load vitamind dataset
vitamind$Z <- rbinom(length(vitamind$X), 2, prob = 0.5) # modify dataset

object <- cosinor.glmm(Y ~ X + amp.acro(time, n_components = 3, group = c("X", NA, "Z"), period = c(12, 10, 8)),
  data = vitamind,
  dispformula = ~ X + amp.acro(time, n_components = 3, group = c("Z", NA, "X"), period = c(12, 11, 8)),
  ziformula = ~ X + amp.acro(time, n_components = 2, group = c("Z", "X"), period = c(12, 8))
)

summary(object)
```

## Using: 'predict.cosinor.glmm.R'

The predict.cosinor.glmm() function allows users to generate predicted
values given a time variable and optional covariates. It is used in the
plot.cosinor.glmm() function to generate the fitted line based on
acosinor.glmm object.

```{r, message=F, warning=F}
data(vitamind)
fit <- cosinor.glmm(Y ~ X + amp.acro(X, group = NA), data = vitamind)
predict(fit, newdata = vitamind)
```

## Plotting cosinor.glmm objects

The GLMMcosinor package includes two ways to visualise cosinor.glmm
objects. Firstly, the function ggplot.cosinor.glmm() creates a
time-response plot of the fitted model:

```{r, message=F, warning=F}
object <- cosinor.glmm(Y ~ X + amp.acro(time, group = "X"), data = vitamind)
plot(object, x_str = "X")
```

This function also allows users to superimpose the datapoints (that the
fit is based on) over the fitted model, using the superimpose.data =
TRUE argument:

```{r, message=F, warning=F}
object <- cosinor.glmm(Y ~ X + amp.acro(time, group = "X"), data = vitamind)
plot(object, x_str = "X", superimpose.data = TRUE)
```

If there are multiple factors in the model, the user can specify which
covariate to be plotted using the x_str argument which accepts a string
corresponding to a group name within the original dataset. By default,
x_str = NULL and the intercept is plotted (all group levels = 0).

The following examples demonstrate how x_str can be used to produce
different plots for the same model;

```{r, message=F, warning=F}
data(vitamind)
vitamind$Z <- rbinom(length(vitamind$X), 3, prob = 0.5)
object <- cosinor.glmm(Y ~ X + amp.acro(time, n_components = 3, group = c("Z", NA, "X"), period = c(12, 10, 8)), data = vitamind)
plot(object)
```

```{r, message=F, warning=F}
data(vitamind)
vitamind$Z <- rbinom(length(vitamind$X), 3, prob = 0.5)
object <- cosinor.glmm(Y ~ X + amp.acro(time, n_components = 3, group = c("Z", NA, "X"), period = c(12, 10, 8)), data = vitamind)
plot(object, x_str = "X")
```

```{r, message=F, warning=F}
data(vitamind)
vitamind$Z <- rbinom(length(vitamind$X), 3, prob = 0.5)
object <- cosinor.glmm(Y ~ X + amp.acro(time, n_components = 3, group = c("Z", NA, "X"), period = c(12, 10, 8)), data = vitamind)
plot(object, x_str = "Z")
```

In addition to time-response plots, the GLMMcosinor package also allows
users to create polar plots. In these plots, the plotted point
represents the acrophase estimate, and the radius represents the
amplitude estimate for a given component. The ellipses represent
confidence regions.

The angle units in the plot can be specified with the 'radial_units'
argument. By default, the units are in radians where a complete
revolution of the plot (2\*pi) represents the maximum period from the
model. The units can be changed to degrees, or even to be expressed in
the same units as the period specification.

By default, the function creates creates polar plots for all components
and stiches them together using the make_cowplot = TRUE argument. If the
user wishes to plot just one component, they can specify this by using
'component_index', though the 'make_cowplot' argument must be FALSE for
this to register.

The direction that the angle increases in can be changed with the
clockwise argument, and the location of the angle = 0 starting point can
be specified with the 'start' argument. Hence, if the user wishes to
create a polar plot that resembles a clock, this can be done by
specifying clockwise = TRUE and start = "top".

The argument: overlay_parameter_info can be used to create a line
extending from the origin to the parameter estimate (to visualise the
amplitude estimate), and a circular arc extending from the angle
starting position (at 0) to the acrophase estimate.

The background grid can also be customised. The argument
'grid_angle_segments' is used to specify how many sectors the polar grid
has, and the 'contour_interval' argument can be used to specify the
distance between adjacent radial circles.

If the user wishes to zoom into the confidence ellipses to show
releveant information, they can adjust the view from the default 'full'
(which plots a full view of the polar plot) to 'zoom' (which enlarges
the smallest view window containing all confidence ellipses), or
'zoom_origin' (which enlarges the smallest view window containing all
confidence ellipses AND the origin).

## Using: 'simulate_cosinor.R'

Simulate_cosinor allows users to simulate circadian data from four
distributions: - family = 'gaussian' - family = 'Gamma(link = 'log')' -
family = 'binomial' - family = 'poisson'

```{r, message=F, warning=F, eval=F}
simulate_cosinor(
  n = 100,
  mesor = 1,
  amp = 2,
  acro = 1.2,
  period = 12,
  n_components = 1,
  beta.group = TRUE,
  beta.mesor = 0.4,
  beta.amp = 0.5,
  beta.acro = 0.2,
  family = c("poisson")
)
```

```{r, message=F, warning=F, echo=F}
df <- simulate_cosinor(
  n = 100,
  mesor = 1,
  amp = 2,
  acro = 1.2,
  period = 12,
  n_components = 1,
  beta.group = TRUE,
  beta.mesor = 0.4,
  beta.amp = 0.5,
  beta.acro = 0.2,
  family = c("poisson")
)

DT::datatable(df)
```

mesor, amp, and acro represent the parameters for the first level, and
the beta.prefix indicates a second level that can be used for data
visualisation (if comparing two groups, for example) if desired.

The simulate_cosinor function can simulate multi-component data. To do
this, specify a vector of values for the parameter inputs (amp and
acro). Note that the mesor term only accepts single-element inputs.
Furthermore, ensure that n_components corresponds to the length of the
vectors: amp, acro, beta.amp, beta.acro, and period.

Here is another example of the simulate_cosinor function. The
cosinor.glmm object has also been included to demonstrate that the
parameter estimates accurately reflect the inputs into the simulate
function:

```{r, message=F, warning=F}
testdata <- simulate_cosinor(1000,
  mesor = 7,
  amp = c(0.1, 0.4, 0.5),
  acro = c(1, 1.5, 0.1),
  beta.mesor = 4.4,
  beta.amp = c(2, 1, 0.4),
  beta.acro = c(1, -1.5, -1),
  family = "poisson",
  period = c(12, 6, 8),
  n_components = 3
)
object <- cosinor.glmm(Y ~ group + amp.acro(times,
  n_components = 3,
  period = c(12, 6, 8),
  group = "group"
), data = testdata, family = poisson())
summary(object)
plot(object, superimpose.data = TRUE, x_str = "group", predict.ribbon = FALSE)
```

##Feature table

```{r, echo = F}
library("flextable")
library("ftExtra")
dat <- dplyr::tibble(
"Package" = c('GLMMCosinor','CosinorPy', 'RhythmCount', 'FMM', 'CircaCompare', 'Cosinor', 'Cosinor2', 'DiscoRhythm', 'LimoRhyde'),
"Language" = c('R','Python','Python', 'R', 'R', 'R', 'R', 'R', 'R'),
"Multi-component" = c('Yes','Yes','Yes','Yes','No','No','No','No','No'), 
"Dispersion models" = c('Yes','No','No','No','No','No','No','No','No'), 
"Zero-inflation models" = c('Yes','No','Yes','No','No','No','No','No','No'), 
"Differential rhythmicity" = c('No','Yes','No','No','No','Yes','Yes','No','Yes'),
"Supported model distributions" = c('Any family listed in glmmTMB (See ?family_glmmTMB)', 
'Gaussian, Poisson, negative-binomial', 
'Poisson, generalised Poisson, zero-inflated Poisson, negative binomial, zero-inflated negative-binomial ', 
'Gaussian','Gaussian','Gaussian','Gaussian','Gaussian','Gaussian'),
"Reference" = c('-','Moskon, 2020','Velikajne, 2022','Fernández, 2022','Parsons,2020', 'Sachs, 2014','Mutak, 2018','Carlucci, 2019','Singer, 2019'))
dat %>%
  flextable::flextable() %>%
  ftExtra::colformat_md() %>%
  flextable::autofit()
```

Vignettes are long form documentation commonly included in packages.
Because they are part of the distribution of the package, they need to
be as compact as possible. The `html_vignette` output type provides a
custom style sheet (and tweaks some options) to ensure that the
resulting html is as small as possible. The `html_vignette` format:

-   Never uses retina figures
-   Has a smaller default figure size
-   Uses a custom CSS stylesheet instead of the default Twitter
    Bootstrap style
