---
title: "Mixed models"
author: "Oliver Jayasinghe and Rex Parsons"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mixed-models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(GLMMcosinor)
library(dplyr)
library(ggplot2)
```

`{GLMMcosinor}` allows specification of mixed models accounting for fixed and/or random effects. Mixed model specification follows the `{lme4}` format. See their vignette, [Fitting Linear Mixed-Effects Models Using lme4](https://cran.r-project.org/package=lme4/vignettes/lmer.pdf), for details about how to specify mixed models.

### Data with subject-level differences

To illustrate an example of using a model with random effects on the cosinor components, we will first simulate some data with `id`-level differences in amplitude and acrophase.

```{r}
f_sample_id <- function(id_num,
                        n = 30,
                        mesor,
                        amp,
                        acro,
                        family = "gaussian",
                        sd = 0.2,
                        period,
                        n_components,
                        beta.group = TRUE) {
  data <- simulate_cosinor(
    n = n,
    mesor = mesor,
    amp = amp,
    acro = acro,
    family = family,
    sd = sd,
    period = period,
    n_components = n_components
  )
  data$subject <- id_num
  data
}

dat_mixed <- do.call(
  "rbind",
  lapply(1:30, function(x) {
    f_sample_id(
      id_num = x,
      mesor = rnorm(1, mean = 0, sd = 1),
      amp = rnorm(1, mean = 3, sd = 0.5),
      acro = rnorm(1, mean = 1.5, sd = 0.2),
      period = 24,
      n_components = 1
    )
  })
)
dat_mixed$subject <- as.factor(dat_mixed$subject)
```

```{r echo=FALSE}
withr::with_seed(
  50,
  {
    dat_mixed <- do.call(
      "rbind",
      lapply(1:30, function(x) {
        f_sample_id(
          id_num = x,
          mesor = rnorm(1, mean = 0, sd = 1),
          amp = rnorm(1, mean = 3, sd = 0.5),
          acro = rnorm(1, mean = 1.5, sd = 0.2),
          period = 24,
          n_components = 1
        )
      })
    )
    dat_mixed$subject <- as.factor(dat_mixed$subject)
  }
)
```


A quick graph shows how there are individual differences in terms of MESOR, amplitude and phase.

```{r}
ggplot(dat_mixed, aes(times, Y, col = subject)) +
  geom_point() +
  geom_line() +
  theme_bw()
```

### A single component model with random effects

For the model, we should include a random effect for the MESOR, amplitude and acrophase as these are clustered within individuals.

In the model formula, we can use the special `amp_acro[n]` which represents the n<sup>th</sup> cosinor component. In this case, we only have one component so we use `amp_acro1`. Following the `{lme4}`-style mixed model formula, we add our random effect for this component and the intercept term (MESOR) clustered within subjects by using `(1 + amp_acro1 | subject)`. The code below fits this model

```r
mixed_mod <- cglmm(
    Y ~ amp_acro(times, n_components = 1, period = 24) + 
      (1 + amp_acro1 | subject),
    data = dat_mixed
  )
```

```{r echo=FALSE}
withr::with_seed(42, {
  mixed_mod <- cglmm(
    Y ~ amp_acro(times, n_components = 1, period = 24) +
      (1 + amp_acro1 | subject),
    data = dat_mixed
  )
})
```

This works by replacing the amp_acro1 with the relevant cosinor components when the data is rearranged and the formula created. The formula created can be accessed using `.$formula`, and shows the `amp_acro1` is replaced by the `main_rrr1` and `main_sss1` (the cosine and sine components of time that also appear in the fixed effects).

```{r}
mixed_mod$formula
```

The mixed model can also be plotted using `autoplot`, but some of the plotting features that are available for fixed-effects models may not be available for mixed-effect models.

```{r}
autoplot(mixed_mod, superimpose.data = TRUE)
```

The summary of the model shows that the input means for MESOR, amplitude and acrophase are similar to what we specified in the simulation (0, 3, and 1.5, respectively).

```{r}
summary(mixed_mod)
```

We can see that the predicted values from the model closely resemble the patterns we see in the input data.

```{r}
ggplot(cbind(dat_mixed, pred = predict(mixed_mod))) +
  geom_point(aes(x = times, y = Y, col = subject)) +
  geom_line(aes(x = times, y = pred, col = subject))
```

This looks like a good model fit for these data. We can highlight the importance of using a mixed model in this situation rather than a fixed effects only model by creating that (bad) model and comparing the two by using the Akaike information criterion using `AIC()`.

```{r}
fixed_effects_mod <- cglmm(
  Y ~ amp_acro(times, n_components = 1, period = 24),
  data = dat_mixed
)

AIC(fixed_effects_mod$fit)
AIC(mixed_mod$fit)
```

Aside from not being able to be useful to see the differences between subjects
from the model, we end up with much worse model fit and likely biased and/or 
imprecise estimates of our fixed effects that we are interested in!

# REML

For mixed models, it may be most appropriate for them to be fit using Restricted
Maximum Likelihood (REML). GLMMcosinor does not automatically use REML for mixed
models but leaves it to the user to specify this when calling `cglmm()`, the 
same as is done by `glmmTMB::glmmTMB()`. Arguments passed to `cglmm(...)` are 
used when calling `glmmTMB::glmmTMB(...)`. This means that the user can control
whether REML is used by using specifying `REML = TRUE`. See `?glmmTMB::glmmTMB`
for more details and what other aspects of the model fitting can be controlled.
Below is the same example as above, but fit using REML, and shows that the model
was fit using REML by inspecting the modelInfo within the glmmTMB fit object.

```r
mixed_mod <- cglmm(
    Y ~ amp_acro(times, n_components = 1, period = 24) + 
      (1 + amp_acro1 | subject),
    data = dat_mixed, REML = TRUE
  )
```

```{r echo=FALSE}
withr::with_seed(42, {
  mixed_mod_reml <- cglmm(
    Y ~ amp_acro(times, n_components = 1, period = 24) +
      (1 + amp_acro1 | subject),
    data = dat_mixed, REML = TRUE
  )
})
```

```{r}
mixed_mod_reml$fit

mixed_mod_reml$fit$modelInfo$REML

mixed_mod$fit$modelInfo$REML
```

# Rhythm estimates at the random-effect level

So far, everything we've looked at (`summary()`, `autoplot()`, `polar_plot()`)
describes the **population-level** (fixed-effect) rhythm. Often we also want
to know the rhythm parameters (amplitude, acrophase, mesor) for each
individual subject - `ranef_estimates()` extracts these by combining each
subject's conditional mode (BLUP) with the population-level estimate.

```{r}
ranef_estimates(mixed_mod)
```

Each component gets its own amplitude/acrophase row per subject; mesor is
reported separately (in a `component = "mesor"` block) since it doesn't vary
by component.

These per-subject estimates can also be overlaid directly on `polar_plot()`
(as points) and `autoplot()` (as fitted curves), via the `ranef_plot` and
`ranef_lines` arguments respectively:

```{r}
polar_plot(mixed_mod, ranef_plot = "subject")
autoplot(mixed_mod, ranef_lines = "subject")
```

## Uncertainty of random-effect estimates

Unlike the population-level estimates from `summary()`, `ranef_estimates()`
doesn't report any uncertainty by default. `glmmTMB` doesn't provide the
joint uncertainty of the conditional modes (random effects) together with
the fixed effects, so the delta-method approach used for the population-level
confidence intervals isn't available here.

Instead, `{GLMMcosinor}` supports a **parametric bootstrap**: simulate new
response data from the fitted model, refit the model to each simulated
dataset, and use the resulting distribution of subject-level estimates to
form percentile confidence intervals. Because this means refitting the model
many times, it can be slow - so it's a deliberate, explicit step
(`add_ranef_boots()`) rather than something that happens automatically. Run
it once, and the result can be reused by every subsequent table or plot that
asks for it, by adding `ranef_ci = TRUE`.

```{r}
# nsim is small here to keep the vignette fast to build - for real analyses,
# prefer the default (nsim = 500) or higher
mixed_mod <- add_ranef_boots(mixed_mod, nsim = 100)
```

`mixed_mod` now carries the bootstrap replicates alongside the rest of the
model, so it can be passed straight into the same functions as before, with
`ranef_ci = TRUE`:

```{r}
ranef_estimates(mixed_mod, ranef_ci = TRUE)
```

This adds `_lower`/`_upper` columns (at the `ci_level` you specify, `0.95`
by default) for amplitude, acrophase, and mesor. The same flag adds a
bootstrap confidence ellipse to `polar_plot()`, and a bootstrap percentile
band to `autoplot()`'s curves:

```{r}
polar_plot(mixed_mod, ranef_plot = "subject", ranef_ci = TRUE)
autoplot(mixed_mod, ranef_lines = "subject", ranef_ci = TRUE)
```

If you try to use `ranef_ci = TRUE` before running `add_ranef_boots()`,
you'll get a message telling you to add it first, e.g.:

```{r, error=TRUE}
mixed_mod_no_boots <- cglmm(
  Y ~ amp_acro(times, n_components = 1, period = 24) +
    (1 + amp_acro1 | subject),
  data = dat_mixed
)
ranef_estimates(mixed_mod_no_boots, ranef_ci = TRUE)
```
